---
title: "Forcing non-endemic singletons: solutions and limitations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Forcing non-endemic singletons: solutions and limitations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(DAISIEprep)
library(ape)
library(phylobase)
```

## Problem statement

When using the "asr" extraction method, non-endemic species that are closely phylogenetically related can be potentially erroneously lumped into a single clade. This clade is labelled endemic as the DAISIE model cannot handle non-endemic clades. For example, see the plot below which shows 3 plant species that are grouped into a single clade even though they are non-endemic island species.

```{r, fig.height=5, fig.width=8}
set.seed(
  1,
  kind = "Mersenne-Twister",
  normal.kind = "Inversion",
  sample.kind = "Rejection"
)
phylo <- ape::rcoal(10)
phylo$tip.label <- c("Plant_a", "Plant_b", "Plant_c", "Plant_d", "Plant_e",
                     "Plant_f", "Plant_g", "Plant_h", "Plant_i", "Plant_j")
phylo <- phylobase::phylo4(phylo)
endemicity_status <- c("not_present", "not_present", "not_present", 
                       "not_present", "not_present", "not_present",
                       "nonendemic", "nonendemic", "nonendemic",
                       "not_present")
phylod <- phylobase::phylo4d(phylo, as.data.frame(endemicity_status))
phylod <- add_asr_node_states(phylod = phylod, asr_method = "mk", rate_model = "ER", tie_preference = "mainland")
plot_phylod(phylod = phylod)
```

If you have _a priori_ knowledge that the non-endemic species should be separate colonisations then this can be taken into account when extracting the phylogenetic island community data using `extract_island_species()`. For example, if you have a group of closely related geographically widespread species, then it is reasonable to assume they did not speciate on the island. 

## First solution: forcing non-endemic island species to be singletons

The `extract_island_species()` function has an argument `force_nonendemic_singletons`, which by default is `FALSE` which fully trusts the ancestral state reconstruction (see `add_asr_node_states()`) allowing non-endemics to be grouped into an endemic clade. **The `force_nonendemic_singletons` argument is only active when using the `extraction_method = "asr"`, if using `extraction_method = "min"` then `force_nonendemic_singletons` will be ignored.** For example:

```{r, extract-with-fns-false}
island_tbl <- extract_island_species(
  phylod = phylod,
  extraction_method = "asr", 
  force_nonendemic_singleton = FALSE
)
island_tbl
```

As you can see the code above extracts only a single colonisation event resulting in an endemic clade.

However, if `force_nonendemic_singletons` is set to `TRUE` then closely related non-endemic island species will be split into single-species island colonists (see Limitations section below for cases when this does not work). For example:

```{r, extract-with-fns-true}
island_tbl <- extract_island_species(
  phylod = phylod,
  extraction_method = "asr", 
  force_nonendemic_singleton = TRUE
)
island_tbl
```

In this case, the `force_nonendemic_singleton = TRUE` now allows non-endemics to be extracted as separate colonisation events to the island.

### Limitation

The first limitation of the approach of setting `force_nonendemic_singleton` to `TRUE` is it does not work when the non-endemic island species are reconstructed to be embedded within an endemic clade. For example, take a look at the tree below and see how the ancestral state reconstruction has a node on the island that is the ancestor of both the endemic and non-endemic species:

```{r, fig.height=5, fig.width=8}
endemicity_status <- c("endemic", "endemic", "endemic", 
                       "endemic", "endemic", "not_present",
                       "nonendemic", "nonendemic", "nonendemic",
                       "not_present")
phylod <- phylobase::phylo4d(phylo, as.data.frame(endemicity_status))
phylod <- add_asr_node_states(phylod = phylod, asr_method = "mk", rate_model = "ER", tie_preference = "mainland")
plot_phylod(phylod = phylod)
```

In this case even when setting `force_nonendemic_singleton = TRUE` the clade is still extracted as a single endemic clade.

```{r, extract-with-fns-true-limitation}
island_tbl <- extract_island_species(
  phylod = phylod,
  extraction_method = "asr", 
  force_nonendemic_singleton = TRUE
)
island_tbl
```

As shown above a **warning** is printed to make you aware that this issue is known and to alert you of possible erroneous results downstream if using this data, for example fitting the DAISIE model.

A second limitation of forcing the non-endemic island species to be separate colonisation events, and in effect overwriting the ancestral state reconstruction, is that the order of the extraction matters. In other words, the order in which `extract_island_species()` goes through the phylogeny extracting the species will influence the extracted island community data. Below we show two examples on the same tree where the order of the endemic and non-endemic species is reversed and how this affects the extraction, the extraction starts from the species labelled "Plant_a". 

```{r, fig.height=5, fig.width=8}
newick <- "(Plant_e:0.9,((Plant_a:0.3,Plant_b:0.3):0.3,(Plant_c:0.3,Plant_d:0.3):0.3):0.3);"
phylo <- ape::read.tree(text = newick)
phylo <- phylobase::phylo4(phylo)
endemicity_status <- c("not_present", "endemic", "endemic", "nonendemic", 
                       "nonendemic")
phylod <- phylobase::phylo4d(phylo, as.data.frame(endemicity_status))
phylod <- add_asr_node_states(phylod = phylod, asr_method = "mk", rate_model = "ER", tie_preference = "mainland")
plot_phylod(phylod = phylod)
island_tbl <- extract_island_species(
  phylod = phylod,
  extraction_method = "asr", 
  force_nonendemic_singleton = TRUE
)
island_tbl
```

When we reverse the order of the endemic and non-endemic clades within the tree, the extraction changes when `force_nonendemic_singleton = TRUE`, even though the ancestral state reconstruction is exactly the same.

```{r, fig.height=5, fig.width=8}
endemicity_status <- c("not_present", "nonendemic", "nonendemic", "endemic", 
                       "endemic")
phylod <- phylobase::phylo4d(phylo, as.data.frame(endemicity_status))
phylod <- add_asr_node_states(phylod = phylod, asr_method = "mk", rate_model = "ER", tie_preference = "mainland")
plot_phylod(phylod = phylod)
island_tbl <- extract_island_species(
  phylod = phylod,
  extraction_method = "asr", 
  force_nonendemic_singleton = TRUE
)
island_tbl
```

This is a known and unwanted issue that the developers of DAISIEprep are working to fix it. For now we recommend to plot the phylogeny with the tip and node states and check if the non-endemic island species need to be separate colonisations (i.e. `force_nonendemic_singleton = TRUE`), if so please be cautious of the influence of extraction order. 

This second issue of extraction order is only applicable when `extraction_method = TRUE` and `force_nonendemic_singleton = TRUE`, when using the `min` extraction algorithm or `force_nonendemic_singleton = FALSE` (the default behaviour) then there is no issue with the order of extraction.

## Second solution: assigning "DAISIE endemics"

In some cases where one or more species from an endemic island radiation have colonised the mainland you may want to reclassify the non-endemic island species (i.e. on the island and the mainland) as an island **endemic**. We term these "DAISIE endemics" as they are not true island endemics, but from the perspective of the DAISIE framework they are the result of speciation on the island and later expanded their range to the mainland. If scoring these non-endemic species as non-endemic it can lead to complex biogeographical reconstructions that may end up breaking up an endemic island radiation into multiple lineages (i.e. multiple island colonisations); and also because this may lead to an underestimation of the speciation rates on the island if the DAISIE model is fitted to the data. By classifying those species as "DAISIE endemics" we can avoid these problems.

```{r, fig.height=5, fig.width=8}
set.seed(
  1,
  kind = "Mersenne-Twister",
  normal.kind = "Inversion",
  sample.kind = "Rejection"
)
phylo <- ape::rcoal(10)
phylo$tip.label <- c("Plant_a", "Plant_b", "Plant_c", "Plant_d", "Plant_e",
                     "Plant_f", "Plant_g", "Plant_h", "Plant_i", "Plant_j")
phylo <- phylobase::phylo4(phylo)
endemicity_status <- c("endemic", "endemic", "endemic", 
                       "endemic", "nonendemic", "endemic",
                       "not_present", "not_present", "not_present",
                       "not_present")
phylod <- phylobase::phylo4d(phylo, as.data.frame(endemicity_status))
phylod <- add_asr_node_states(phylod = phylod, asr_method = "mk", rate_model = "ER", tie_preference = "mainland")
plot_phylod(phylod = phylod)
```

By assinging species that were likely endemic before recolonising the mainland or expanding their range to another region as _endemic_ it resolves issues of incorrectly breaking apart island radiations when using the `min` extraction method, or creating overly complex reconstructions when using `add_asr_node_states()` and the `asr` extraction method (although this latter `asr` approach is less vulnerable to erroneous reconstructions in many of these back-colonisation scenarios).

The example below is the same phylogeny but the "Plant_e" species is now reclassified as endemic ("DAISIE endemic") for the purposes of reconstruction and extraction as it is judged by the empiricist to be the result of speciation on the island, and thus from the perspective of evolutionary processes on the island it is endemic. 

```{r, fig.height=5, fig.width=8}
endemicity_status <- c("endemic", "endemic", "endemic", 
                       "endemic", "endemic", "endemic",
                       "not_present", "not_present", "not_present",
                       "not_present")
phylod <- phylobase::phylo4d(phylo, as.data.frame(endemicity_status))
phylod <- add_asr_node_states(phylod = phylod, asr_method = "mk", rate_model = "ER", tie_preference = "mainland")
plot_phylod(phylod = phylod)
```

In the two cases shown above using the `asr` extraction would have resulted in the same island community data, but when using larger, more complex phylogenetic and biogeographic data (e.g. birds of Madagascar) using the "DAISIE endemics" approach can help simplify data processing.

### Limitations

This approach is only valid if you know the island system well and are sure that the non-endemic species within an island radiation are most likely the result of island speciation. By augmenting the endemicity status of the island species in the way outlined above it introduce biological inaccuracies if not done with care. 

In the case of the issue of having non-endemics that you believe to be _true non-endemmic island species_, and these are embedded within an island radiation of island endemic species, then this issue outlined at the start of this vignette remains and neither reclassifying nor using `force_nonendemic_singletons = TRUE` will resolve the issue.

## Solution are not mutually exclusive

The two solutions covered in this article can be used jointly. For example if you have an endemic radiation with a single non-endemic and a group of closely related non-endemic species in a different part of the phylogeny then the endemicity status can be updated and `force_nonendemic_singleton = TRUE` when calling `extract_island_species()`. For example: 

```{r, fig.height=5, fig.width=8}
set.seed(
  1,
  kind = "Mersenne-Twister",
  normal.kind = "Inversion",
  sample.kind = "Rejection"
)
phylo <- ape::rcoal(10)
phylo$tip.label <- c("Plant_a", "Plant_b", "Plant_c", "Plant_d", "Plant_e",
                     "Plant_f", "Plant_g", "Plant_h", "Plant_i", "Plant_j")
phylo <- phylobase::phylo4(phylo)
endemicity_status <- c("not_present", "nonendemic", "endemic", 
                       "endemic", "endemic", "not_present",
                       "not_present", "nonendemic", "nonendemic",
                       "not_present")
phylod <- phylobase::phylo4d(phylo, as.data.frame(endemicity_status))
phylod <- add_asr_node_states(phylod = phylod, asr_method = "mk", rate_model = "ER", tie_preference = "mainland")
plot_phylod(phylod = phylod)
island_tbl <- extract_island_species(
  phylod = phylod,
  extraction_method = "asr"
)
island_tbl
```

In this example the island community data extracted is two endemic clades (by default in `extract_island_species()` the `force_nonendemic_singleton = FALSE`).

If we set the non-endemic species in the island radiation to _endemic_ using the "DAISIE endemic" reasoning stated above, and by manually setting `force_nonendemic_singleton = TRUE` we can acheive, for particular scenarios, be better extraction.

```{r, fig.height=5, fig.width=8}
endemicity_status <- c("not_present", "endemic", "endemic", 
                       "endemic", "endemic", "not_present",
                       "not_present", "nonendemic", "nonendemic",
                       "not_present")
phylod <- phylobase::phylo4d(phylo, as.data.frame(endemicity_status))
phylod <- add_asr_node_states(phylod = phylod, asr_method = "mk", rate_model = "ER", tie_preference = "mainland")
plot_phylod(phylod = phylod)
island_tbl <- extract_island_species(
  phylod = phylod,
  extraction_method = "asr",
  force_nonendemic_singleton = TRUE
)
island_tbl
```

## DAISIEprep development

The DAISIEprep team are actively working on the issues outlined in this vignette and hope to make improvements over the next few months. Please keep an eye on new versions of the DAISIEprep package being released in case improvements become available.

