---
title: "Performance"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Performance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(DAISIEprep)
library(ape)
if (requireNamespace("microbenchmark", quietly = TRUE)) {
  library(microbenchmark)
} else {
  stop("Microbenchmark package needed for this vignette")
}
```

First, we explain the setup of the performance analysis and then loop over this
to run the analysis.

```{r create parameter space}
parameter_space <- expand.grid(
  tree_size = c(10, 100, 1000, 10000),
  prob_on_island = c(0.2, 0.5),
  prob_endemic = c(0.2, 0.8),
  sampling_type = c("uniform", "clustered")
)
```

```{r tree simulation} 
set.seed(
  1,
  kind = "Mersenne-Twister",
  normal.kind = "Inversion",
  sample.kind = "Rejection"
)
phylo <- ape::rcoal(n = parameter_space$tree_size[1])

# generate a set of unique tip labels that conform to standard
tip_labels <- expand.grid(letters, letters, letters)
tip_labels <- do.call(paste0, tip_labels)
tip_labels <- tip_labels[1:parameter_space$tree_size[1]]
tip_labels <- paste("bird", tip_labels, sep = "_")
phylo$tip.label <- tip_labels
```

```{r endemicity simulation}
prob_not_present <- 1 - parameter_space$prob_on_island[1]
prob_endemic <- 
  parameter_space$prob_endemic[1] * parameter_space$prob_on_island[1]
prob_nonendemic <- 
  (1 - parameter_space$prob_endemic[1]) * parameter_space$prob_on_island[1]

# uniform sampling
endemicity_status <- sample(
  x = c("endemic", "nonendemic", "not_present"),
  size = parameter_space$tree_size[1],
  replace = TRUE,
  prob = c(prob_endemic, prob_nonendemic, prob_not_present)
)

# clustered sampling
#TODO
```

As we are stochastically simulating the endemicity statuses of the species in 
the tree it may be that there is not an outgroup species not on the island in
order to correctly extract the colonisation times from the stem age of the 
species or clade on the island. Therefore, we add an outgroup which we ensure
is not present on the island.

```{r add an outgroup}
phylo <- add_outgroup(phylo)
endemicity_status <- c("not_present", endemicity_status)
```

```{r format data for DAISIEprep}
phylod <- phylobase::phylo4d(phylo, as.data.frame(endemicity_status))
phylod <- add_asr_node_states(
  phylod = phylod,
  asr_method = "parsimony",
  tie_preference = "island",
  earliest_col = FALSE
)
```

```{r run extraction}
time <- microbenchmark::microbenchmark(
  island_tbl <- extract_island_species(
    phylod = phylod,
    extraction_method = "min",
    island_tbl = NULL,
    include_not_present = FALSE
  ),
  times = 10L
)
```

```{r median run time}
median_time <- median(time$time)
```

```{r run performance analysis, eval=FALSE}

times_list <- list()
for (i in seq_len(nrow(parameter_space))) {
  times <- c()
  for (j in 1:10) {
    
    # simulate phylogeny
    phylo <- ape::rcoal(n = parameter_space$tree_size[i])
    
    
    # generate a set of unique tip labels that conform to standard
    tip_labels <- expand.grid(letters, letters, letters)
    tip_labels <- do.call(paste0, tip_labels)
    tip_labels <- tip_labels[1:parameter_space$tree_size[i]]
    tip_labels <- paste("bird", tip_labels, sep = "_")
    phylo$tip.label <- tip_labels
    prob_not_present <- 1 - parameter_space$prob_on_island[i]
    prob_endemic <- parameter_space$prob_endemic[i]
    prob_nonendemic <- 1 - parameter_space$prob_endemic[i]
    
    if (parameter_space$sampling_type[i] == "uniform") {
      
      empty_island <- TRUE
      while (empty_island) {
        # generate tip states under uniform sampling
        endemicity_status <- sample(
          x = c("endemic", "nonendemic", "not_present"),
          size = parameter_space$tree_size[i],
          replace = TRUE,
          prob = c(prob_endemic, prob_nonendemic, prob_not_present)
        )
        if (any(endemicity_status != "not_present")) {
          empty_island <- FALSE
        }
      }
    } else if (parameter_space$sampling_type[i] == "clustered") {
      # clustered sampling
      #TODO
    }
    
    # add not present outgroup
    phylo <- add_outgroup(phylo)
    endemicity_status <- c("not_present", endemicity_status)
    
    # format data for DAISIEprep
    phylod <- phylobase::phylo4d(phylo, as.data.frame(endemicity_status))
    phylod <- add_asr_node_states(
      phylod = phylod,
      asr_method = "parsimony",
      tie_preference = "island",
      earliest_col = FALSE
    )
    
    # run extraction
    time <- microbenchmark::microbenchmark(
      island_tbl <- extract_island_species(
        phylod = phylod,
        extraction_method = "min",
        island_tbl = NULL,
        include_not_present = FALSE
      ),
      times = 10L
    )
    times[j] <- median(time$time)
  }
  times_list[[i]] <- times
}
```
