---
title: "Sensitivity"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Sensitivity}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center", 
  fig.height = 5, 
  fig.width = 7
)
```

## Introduction

In this vignette we provide a set of best practices for sensitivity analyses 
that should be taken into consideration when conducting data formatting and 
model fitting in the DAISIE and DAISIEprep framework. Sensitivty analyses is the
testing of a model output to variations in the model input. A model is considered
sensitive if the model output vastly changes due to relatively small 
perturbations in the data input into the model. These small perturbations may be
due to uncertainty in the data (i.e. a posterior distribution of possible 
branching times in a phylogeny), measurement error, or other factors. In the 
case we are interested in, these perturbations are variation in island 
colonisation and branching times, endemicity status on the island and number
of colonisations events. Each of these can change if using multiple phylogenies 
from the posterior distribution of inferred phylogenies, or using different 
extraction algorithms in DAISIEprep. 

Firstly we discuss thetype of variations in the data and how each of these can 
impact parameter estimation. Another aspect of sensitivity analysis, which we 
will not explore in this vignette, is the sensitivity of model selection to 
input data variation. By this we mean the best-fit model and the ranking of the 
model (by likelihood, AIC, BIC, or other metric) may change given differences in
the data. This can be equally as important as sensitivity of parameter estimates 
and we recommend users check model selection as well as parameter estimates in 
their work.

## Sensitivity to extraction method

```{r setup}
library(DAISIEprep)
```

The `DAISIEprep` R package provides the tools to extract phylogenetic community
data from phylogenetic trees with the endemicity status of the species assigned
to each tip in the phylogeny. However, there is not a single correct method for
extracting the data and thus `DAISIEprep` implements several algorithms to 
account for variations in what would be considered an appropriate assumptions
for the island system of interest. The two major divisions for extracting data
are in the `extraction_method` argument in `extract_island_species()` (and
by extension `extract_multi_island_species()`), which can be either `"min"` for 
the minimum time of colonisation algorithm, or `"asr"` for the geographical 
ancestral state reconstruction algorithm. The `"min"` algorithm conforms to the 
assumptions of the DAISIE inference model (implemented in the [`DAISIE` R package](https://cran.r-project.org/web/packages/DAISIE/index.html)). These 
assumptions are: 

  * Non-endemic species cannot be part of an island clade
  * There is no back-colonisation from the island away to the mainland or other
  island 
  * There cannot be a species not present on the island nested in an endemic 
  island clade
  
The all points are linked, by not allowing back-colonisation, a species on the islands 
cannot be endemic to the island (i.e. part of an endemic island radiation) and
then migrate or expand its range away from the island. These three points mean
that if the island system of interest has experienced back-colonisation or a
species in an endemic island radiation has expanded its range off the island and
its island population has gone extinct (making it not present on the island but
extant) the `"min"` algorithm will split clades into multiple colonisations. In 
the case that the island of interest is very remote and species colonisation and 
diversify and do not disperse off the island, this algorithm provides a simple 
model to extract the data.

However, it is clear that this common assumptions of the DAISIE model and 
thus the `"min"` algorithm are violated in empirical data. Therefore the second
algorithm, `"asr"`, is implemented to remedy this. The 
`"asr"` algorithm uses the reconstructed states at each node in the phylogeny inferring 
whether a species is not present on the island, non-endemic to the island, or
endemic to the island. Using this information the algorithm can traverse the
phylogeny back to the node where the island clade colonised the island. This
algorithm overcomes the limitations of the `"min"` algorithm by allowing 
non-endemic species to be part of island clades (extracting them as endemic
clades for the purposes of applying the DAISIE inference model), and additionally
allowing species that are not present on the island to be included in data when
embedded within an endemic island clade (this feature is turned on/off with the
`include_not_present` argument in `extract_island_species()`). Therefore, the 
`"asr"` algorithm has benefits when the focal island system has experienced some
species movement from the island to other regions. However, it is not without
limitations, ancestral state reconstruction models should be interpreted with
caution and uncertainty of a species geographic range deep in the past, near the
root of the tree is often high leading to variability in interpretation of 
whether a species was present on the island at the time. The formulation of the 
ancestral state reconstruction model is also important, with the transition 
matrix between states crucial to plausible results. By default we use a 
symmetrical transition structure where species go from not present on the island, 
to non-endemic and then to endemic. Without jumps from not present to endemic 
and vice versa. This is in line with the reasoning in the DAISIE model that 
species that colonise the island do not migrate their entire mainland 
population, instead going through a widespread range, before becoming endemic
via cladogenesis or anagenesis on the island, or extinction of the mainland
population.

In this vignette we demonstrate the sensitivity of the parameters estimated
from the DAISIE maximum likelihood inference model to changes in the algorithm 
used to extract the data. We apply the `"min"` and `"asr"` algorithms, and within 
`"asr"` we apply two different models of ancestral state reconstruction: parsimony
and continuous-time Markov model (Mk model). Traditionally, these have been two
of the most common methods for reconstructing ancestral states, for other 
methods to reconstruct ancestral ranges see Extending_asr vignette in the 
`DAISIEprep` package.

The data we use for this example is the macro-phylogeny of mammals (Upham et 
al., 2019) and the island endemicity data of Madagascar (Michielsen et al., 2021). 
This is a large data set containing ...

The results presented in this vignette are not computing each time the vignette
is rendered due to the large computation time required. Instead, the analyses are
run on a cluster computer and saved in the package. The analysis script run to 
produce the results can be found in the [DAISIEprepExtra](https://github.com/joshwlambert/DAISIEprepExtra)
package [here](https://github.com/joshwlambert/DAISIEprepExtra/tree/main/inst/scripts).
The sensitivity analysis uses the `sensitivity()` function in the `DAISIEprep` 
package. 

The `sensitivity()` function creates a table of all possible combination of
data extraction settings, given the input arguments, and this forms our parameter 
space for the sensitivity analysis. The phylogenetic trees and island endemicity
data is provided to carry out the extraction and formatting. `DAISIEprep` uses
the `phylo4d` class from the `phylobase` R package. However, for the `sensitivity()`
function a `phylo` object can be provided and all the house-keeping is taken care
of inside the function. The `sensitivity()` function loops through each parameter
setting and extracts and formats the data and fit the DAISIE
maximum likelihood inference model (`DAISIE::DAISIE_ML_CS()`) to the data.

The output produces results for the DNA-only phylogeny and the complete
phylogeny. The raw data of parameter estimates for the different parameter 
settings is tidied into a tibble containing the data we need for both the DNA
and complete phylogeny. 

```{r, echo=FALSE}
sensitivity_data <- DAISIEprep:::read_sensitivity()
```

We can plot the distribution of parameter estimates
for the DNA and complete data sets.

```{r, eval=FALSE}
DAISIEprep:::plot_sensitivity(sensitivity_data = sensitivity_data$dna_data)
``` 

```{r, eval=FALSE}
DAISIEprep:::plot_sensitivity(sensitivity_data = sensitivity_data$complete_data)
``` 
