---
title: "Tutorial"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(DAISIEprep)
library(ape)
library(phylobase)
library(ggtree)
```

First we simulate a phylogeny using the package `ape`

```{r simulate phylogeny}
set.seed(1)
phylo <- ape::rcoal(10)
```

```{r plot phylogeny}
ape::plot.phylo(phylo)
```

Then we convert to the phylogeny to a `phylo4` class defined in the package
`phylobase`. This allows use to easily work species data for each tip in the 
phylogeny, for example island endemicity status.

```{r convert phylo to phylo4}
phylo <- as(phylo, "phylo4")
phylobase::plot(phylo)
```

Now we have a phylogeny in the `phylo4` format we can easily append data. First
we randomly simulate island endemicity patterns, assuming each species has an
equal probability of being not on the island (`"not_present"`), endemic to the
island (`"endemic"`) or non-endemic to the island (`"nonendemic"`).

```{r create island endemcity data}
endemicity_status <- sample(c("not_present", "endemic", "nonendemic"), 
                            size = length(phylobase::tipLabels(phylo)), 
                            replace = TRUE)
```

Next we can add the endemicity data to our phylogenetic data using the `phylo4d`
class, again from the `phylobase` package. This call is designed for
phylogenetic and trait data to be stored together. The endemicity status needs
to be converted to a data frame in order for the column to be labelled 
correctly.

```{r convert to phylo4d}
phylod <- phylobase::phylo4d(phylo, as.data.frame(endemicity_status))
```

We can now visualise our phylogeny with the island endemicity status plotted at
the tips. This uses the `ggtree` and `ggplot2` packages.

```{r plot phylogeny with tip data}
ggtree::ggtree(phylod) +
  ggtree::geom_tippoint(
    ggplot2::aes(colour = endemicity_status), 
    size = 3
  )
```

Now that we can see the tips that are present on the island we can start to 
extract them to form our island community data set that can be used in the
`DAISIE` R package to fit likelihood models of island colonisation and 
diversification.

Before we extract species we will first create an object to store all of the 
island colonist's information. This uses the `island_tbl` class introduced
in this package (`DAISIEprep`). The `island_tbl` is an S4 class that holds
a data. This `island_tbl` object can then easily be converted to a
`DAISIE_datatable` using the function `as_daisie_datatable` (more information
on this to below).

```{r create island_tbl, eval=FALSE}
island_tbl <- methods::new("island_tbl")
island_tbl
```

We can see that this is a class containing an empty data frame. In order to fill
this data frame with information on the island colonisation and diversification
events we can run:

```{r extract island data, eval=FALSE}
island_tbl <- extract_island_species(phylod = phylod)
```

To see an example of the DAISIE_datatable object see

```{r Galapagos data table, eval=FALSE}
data("Galapagos_datatable", package = "DAISIE")
```

Each row in the table represents an independent colonisation event. The table has four columns:

* `Clade_name`: name of the independent colonisation event.
* `Status`: One of the following categories:
  + `Endemic`: applicable for both anagenetic species and radiations.
  + `Non_endemic`: if the taxon is not endemic to the island, and the age of the colonisation is based on a phylogeny where both island and non-island populations of the species have been sampled.
  + `Non_endemic_MaxAge`: if the taxon is not endemic to the island, and only an upper bound to the time of colonisation of the island is known. This applies if individuals from the island population of the species have not been sampled, but an age of the species is known. 
  + `Endemic&Non_Endemic` when an endemic clade is present and the mainland ancestor has re-colonised. For remote islands this is expected to be very rare. 
* `Missing_species`: number of island species that were not sampled for a particular clade (only applicable for radiations).
* `Branching_times`: this should be the stem age of the population/species in the case of `Non_endemic`, `Non-endemic_MaxAge` and `Endemic` anagenetic species. For cladogenetic species these should be branching times of the radiation including the stem age of the radiation. Note - if there are species within the radiation that are not found on the island (e.g. back-colonisation) the branching times of these species should be excluded, as the mainland species pool is treated as static.
