---
title: "Tutorial"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(DAISIEprep)
library(ape)
library(phylobase)
library(ggtree)
```

First we simulate a phylogeny using the package `ape`

```{r simulate phylogeny}
set.seed(1)
phylo <- ape::rcoal(10)
```

```{r plot phylogeny}
ape::plot.phylo(phylo)
```

Then we convert to the phylogeny to a `phylo4` class defined in the package
`phylobase`. This allows use to easily work species data for each tip in the 
phylogeny, for example island endemicity status.

```{r convert phylo to phylo4}
phylo <- as(phylo, "phylo4")
phylobase::plot(phylo)
```

Now we have a phylogeny in the `phylo4` format we can easily append data. First
we randomly simulate island endemicity patterns, assuming each species has an
equal probability of being not on the island (`"not_present"`), endemic to the
island (`"endemic"`) or non-endemic to the island (`"nonendemic"`).

```{r create island endemcity data}
endemicity_status <- sample(c("not_present", "endemic", "nonendemic"), 
                            size = length(phylobase::tipLabels(phylo)), 
                            replace = TRUE)
```

Next we can add the endemicity data to our phylogenetic data using the `phylo4d`
class, again from the `phylobase` package. This call is designed for
phylogenetic and trait data to be stored together. The endemicity status needs
to be converted to a data frame in order for the column to be labelled 
correctly.

```{r convert to phylo4d}
phylod <- phylobase::phylo4d(phylo, as.data.frame(endemicity_status))
```

We can now visualise our phylogeny with the island endemicity status plotted at
the tips. This uses the `ggtree` and `ggplot2` packages.

```{r plot phylogeny with tip data}
ggtree::ggtree(phylod) +
  ggtree::geom_tippoint(
    ggplot2::aes(colour = endemicity_status), 
    size = 3
  )
```

Now that we can see the tips that are present on the island we can start to 
extract them to form our island community data set that can be used in the
`DAISIE` R package to fit likelihood models of island colonisation and 
diversification.

Before we extract species we will first create an object to store all of the 
island colonist's information. This uses the `daisie_data_tbl` class introduced
in this package (`DAISIEprep`). The `daisie_data_tbl` is an S4 class that holds
a data

